---
title: "Ranking of Climatalogical Storm Events - Harm to Human Health and Economic Consequence"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

This analysis uses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

To support decisions about prioritization of resources to be used for storm preparedness, this analysis ranks the human health and economic consequences of various types of events.  

In the human health area, tornadoes are by far the most harmful storm events, with a total of 220 categories reported. The largest economic impact has been caused by flood, with 429 categories reported. 

## Data Processing
Libraries used in this analysis are loaded here.  

```{r message = FALSE}
library(dplyr)
library(tidyr)
```
The original bzip2 compressed data file was downloaded using the link:

<https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2> 

The file was placed in the project directory and is the starting point for the following processing operations.

#### Loading the Raw Data into R

The R function `read.csv` is able to read data directly from the compressed version of the csv file. 

```{r raw.data, cache = TRUE}
raw.data <- read.csv(file = "repdata_data_StormData.csv.bz2")
```

#### Characterizing the data

There are 902297 rows and 37 columns in this data.

```{r dim}
dim(raw.data)
names(raw.data)
```

#### Data Cleaning and Transformation

Many of the 37 columns in the data do not contribute to this analysis so they are removed here to help with efficient execution.  

```{r clean, cache=TRUE}
clean.data <- raw.data %>% select(REFNUM, EVTYPE, BGN_DATE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

The following column type transformations are handled next.  
1. Because the BGN_DATE column is initially read as a factor variable, we convert to a date type. 
2. Columns FATALITIES and INJURIES are converted from numeric to integer.  

```{r columntypes, cache=TRUE}
clean.data <- clean.data %>% 
  mutate(BGN_DATE = as.Date(BGN_DATE, format = "%m/%d/%Y"),
         FATALITIES = as.integer(FATALITIES),
         INJURIES = as.integer(INJURIES)) 
```

Each variable relating to a category of economic cost is represented in the raw data by 2 columns, which must be combined to obtain the actual monetary value.  For example, to obtain the value of property damage, the numeric value of the PROPDMG column must be adjusted by a multiplier that is related to values of the PROPDMGEXP field.  Crop damage is evaluated by similarly combining the CROPDMG and CROPDMGEXP columns.  

The multiplier is inferred from the single character values that follow:

```{r EXPList}
EXPchars <- union(clean.data$PROPDMGEXP, clean.data$CROPDMGEXP)
print(EXPchars)
```

The correct interpretation of these factor values is not immediately obvious.  An attempt to define these value semantics is documented publicly at (https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html). However, the conclusion from that document that numeric values in the range of 0 to 8 all imply a fixed, constant multiplier of 10 seems unlikely.  Looking at all property damage values for which the PROPDMGEXP is 5 or higher, the nonzero PROPDMG values range from 0.1 to 88.0. Using the proposed multiplier of 10 for measurements results in overall economic costs in the range of 1 to 880 dollars due to storm events.  This can't possibly be realistic.  

More likely would be that the value is intended to be applied as an exponent of 10 to obtain the multiplier.  For the same 

```{r multipliers, cache=TRUE}
interpret.EXP <- function(mantissa, exponent) {
  return = numeric(length = length(mantissa))
  for (index in 1:length(mantissa)) {
    mant <- as.character(mantissa[index])
    exp <- as.character(exponent[index])
    val.string <- switch(EXPR = toupper(exp),
                         H=paste0(mant,"e2"),
                         K=paste0(mant,"e3"),
                         M=paste0(mant,"e6"),
                         B=paste0(mant,"e9"),
                         "1"=, "2"=, "3"=, "4"=, "5"=, "6"=, "7"=, "8"= paste0(mant,"e",exp),
                         "+"=1,
                         "?"=, "-"=, "0")
    return[index] <- as.numeric(val.string)
  }
  return
}

clean.data <- clean.data %>% 
  mutate(Property.Damage = interpret.EXP(PROPDMG,PROPDMGEXP), 
         Crop.Damage = interpret.EXP(CROPDMG,CROPDMGEXP)) %>%
  select(-c(PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP))
```

The following histogram shows counts of all measurements in the raw data, grouped by decade.  It is evident that data collection efforts have yielded a dramatically increasing volume of recorded storm data over the years. 

```{R}
hist(clean.data$BGN_DATE, 
     breaks = as.Date(c("19500101", "19600101", "19700101", "19800101", "19900101", "20000101", "20100101", format(max(clean.data$BGN_DATE),"%Y%m%d")), "%Y%m%d"), 
     freq = TRUE,
     xlab = "Decade",
     main = "Numbers of Recorded Storm Events by Decade")
```

## Results

Grouping the data by event type allows the analysis to examine aggregated statistics for each type of event. 

```{R}
grouped <- clean.data %>%
  group_by(EVTYPE)
```

#### Storm Related Harm to Human Health

The following list depicts the storm events causing harm to human health, in descending order of total people impacted. 

```{R}
health <- grouped %>% 
  summarise(Deaths = sum(FATALITIES), Injuries = sum(INJURIES)) %>% 
  mutate(Total = Deaths + Injuries) %>%
  filter(Total > 0) %>%
  arrange(desc(Total))
as.data.frame(health)
```

#### Evaluating Economic Consequences

The following list depicts the storm events causing economic consequences, in descending order of total cost (US$).  

```{R}
options(width = 80)
econ <- grouped %>% 
  summarise(Property.Damage = sum(Property.Damage), Crop.Damage = sum(Crop.Damage)) %>% 
  mutate(Total = Property.Damage + Crop.Damage) %>%
  filter(Total > 0) %>%
  arrange(desc(Total))
as.data.frame(econ)
```

